Robertson and Cao Project

### 2024-02-22 code/eda-code/EDA-code.qmd
Added to EDA, staged in Manuscript
I added 5 figures/ tables to the EDA and in addition I added the EDA chunk to the manuscript document. I would prefer to just add the figure images preceding the description, but I cannot get markdown to allow me to pull just the figures.

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./R/eda-code/edacode.R"))
```

@@

The following data set has been cleaned. Any observations with NAs has been removed. 
```{r, table1}

@@

## Addressing Numerical Data Set
The distribution of the numerical income to poverty ratio can be shown below. It displays a value from 1-3. A ratio displaying less than 1 depicts an income less than the poverty level. Any income ratio >1 indicates an income greater than the poverty level. Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.

@@

The family income variable provides a more normal distribution with the majority of the observation coming from $75000 income level.

@@

The race and ethnicity variables show a normal distribution.
```{r, RACEETHK}

@@

The Insurance status variable shows that the majority of the observations came from households with private insurance only. This is mirrored in the facility variable, where a majority of its observation s show private facilities being the location.

@@

## Model and Plot Fitting

@@

### Fit 2
A boxplot is created with income-poverty ratio to insurance status. The box plot indicates there are distinct difference between the two status. A statistical model is fitted to the same variables. The p-value of < 2.26e-16 suggests there is statistical significance between the income-poverty variable and the insurance status.

@@

>>>>

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./code/eda-code/edacode.R"))
```

@@

The following data set has been cleaned.
Any observations with NAs has been removed.

```{r, table1}

@@

## Addressing Numerical Data Set

The distribution of the numerical income to poverty ratio can be shown below.
It displays a value from 1-3.
A ratio displaying less than 1 depicts an income less than the poverty level.
Any income ratio \>1 indicates an income greater than the poverty level.
Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.

@@

The family income variable provides a more normal distribution with the majority of the observation coming from \$75000+ income level.

@@

The race and ethnicity variables show a distribution that is similar to the U.S. census data, which concludes that the racial composition of the U.S. is 58.9% white, 13.6% African American, and 19.1% Hispanic or Latino.

```{r, RACEETHK}

@@

The Insurance status variable shows that the majority of the observations came from households with private insurance only. This is followed by any medicaid and then other insurance.
This is mirrored in the facility variable, where a majority of its observations show private facilities being the location.

@@

We examine the percentage of up-to-date vaccination status by state.
```{r, state-vaccine}

```{r, facility-vaccine}

```{r, race-ethnicity-vaccine}

```
## Model and Plot Fitting

@@

### Fit 2

A boxplot is created with income-poverty ratio to insurance status.
The box plot indicates there are distinct difference between the two status.
A statistical model is fitted to the same variables.
The p-value of \< 2.26e-16 suggests there is statistical significance between the income-poverty variable and the insurance status.

@@

### Fit 3
```{r, fitfig3}

### Fit 4
```{r, fitfig4}

### 2024-02-22 code/processing-code/Processing-Quarto-Output.qmd
Added to EDA, staged in Manuscript
I added 5 figures/ tables to the EDA and in addition I added the EDA chunk to the manuscript document. I would prefer to just add the figure images preceding the description, but I cannot get markdown to allow me to pull just the figures.

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./R/processing-code/creating-df/Creating_Data_Frame.R"))
```

@@

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so: package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives", but it adds typing.
You can do it either way.

@@

By inspecting the data as done above, we find some problems that need addressing:

First, there is an entry for height which says "sixty" instead of a number.
Does that mean it should be a numeric 60?
It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
Since we don't know how to fix this, we might decide to remove the person.
This "sixty" entry also turned all Height entries into characters instead of numeric.
That conversion to character also means that our summary function isn't very meaningful.
So let's fix that first.

@@

Now we see that there is one person with a height of 6.
That could be a typo, or someone mistakenly entered their height in feet.
Since we unfortunately don't know, we might need to remove this person, which we'll do here.

@@

All done, data is clean now.
Time to save.
Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

Finally, we save the clean data as RDS file.
I suggest you save your processed and cleaned data as RDS or RDA/Rdata files.
This preserves coding like factors, characters, numeric, etc.
If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text.
If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data: http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

@@

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./R/processing-code/cleaningdata/cleaning1.R"))
```

>>>>

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./code/processing-code/creating-df/Creating_Data_Frame.R"))
```

@@

@@

@@

@@

@@

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./code/processing-code/cleaningdata/cleaning1.R"))
```

### 2024-02-22 products/manuscript/Manuscript.qmd
Added to EDA, staged in Manuscript
I added 5 figures/ tables to the EDA and in addition I added the EDA chunk to the manuscript document. I would prefer to just add the figure images preceding the description, but I cannot get markdown to allow me to pull just the figures.

---
title: "Robertson and Cao Project"
subtitle: "Teenage HPV Vaccination Coverage and Socioeconomic Factors"
author: Kelly Cao and Rachel Robertson
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../assets/dataanalysis-references.bib
csl: ../assets/apa.csl
---
# *Part 1*
## Data Set
  The data we chose to use is the National Immunization Survey (NIS) of 2022. It consists survey data that was collected to monitor vaccination coverage for teenagers age 13-17 in the United States of America. The surveys were conducted by the National Center for Immunization and Respiratory Diseases of the Center (NCIRD) for Disease Control and Prevention (CDC). The survey itself consists of two parts: (1) the household telephone (random digit dialing) survey answered by a parent or guardian and (2) a mailed survey for the vaccination provider, called “Immunization History Questionnaire.”
  The original data set is a fixed width file (FWF). In the data set, there are 43,738 rows and 672 columns, with each row representing an individual and the each column representing the answer to a question. The questions pertain to the teenager’s immunization history, demographics, and additional household-reported health information. 
We have truncated the data set to include 26-selected columns for the ease of observation. We have also filtered the data set to include one year, 2022, which is the most recent year of data that is published. The provided survey data will require cleaning, as there are signs of human errors present. 
## Opening the Data Set

First, we will load the readr package to open the file
```{r}
library(readr) #Loading the package to read the fwf
```

Next, we will define the column widths (beginning and end positions) and applying the column names to each position.
```{r}
column_position<- fwf_positions(
  start = c(1, 89, 287, 295, 298, 314, 330, 331, 334, 336, 343, 344, 345,
            347, 355, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
  ),
  end = c(5, 92, 288, 295, 313, 329, 330, 332, 334, 336, 343, 344, 346,
          347, 356, 359, 362, 472, 473, 495, 496, 497, 502, 504, 1296, 1297
  ),
  col_names = c("SEQNUMT", "YEAR", "AGE", "EDUC1", "INCPORAR", "INCPORAR_I", 
                "INCPOV1", "INCQ298A", "LANGUAGE", "MOBIL_1", "RACEETHK", 
                "RACE_K", "RENT_OWN", "SEX", "STATE", "FACILITY", "WELLCHILD", 
                "P_U13HPV", "P_U13HPV3", "P_UTDHPV", "P_UTDHPV_15", 
                "P_UTDHPV_15INT", "P_UTDHPV2", "P_UTDHPV3", "INS_STAT2_I", 
                "INS_BREAK_I")
)

@@

Create the data frame "data" with read_fwf function. 
```{r}
data_frame <- read_fwf( "../products/NISTEENPUF22.DAT", 
                  col_positions = column_position, 
                  col_types = cols(.default = "c")
)
```

```{r}
#Read the statistics in the data file.
  summary(data_frame)
  str(data_frame)
```

The data will then be cleaned. As of now, the data set requires a library to translate the code.

The data will then be converted the data frame to a readable and downloadable .csv file for further analysis
```{r}
output_file<-"NIS_Teen_Data_2022.csv"
write.csv(data_frame, file = output_file, row.names = FALSE)
```

## Research Question
  We want to use the 2022 National Immunization Survey for teens to identify which socioeconomic and geographic factors are associated with HPV vaccination completion for teenagers in the U.S. Specifically, we are asking the question, What is the likelihood of a teenager being up-to-date for their HPV vaccinations in the U.S., based on socioeconomic factors and geographic distribution. The outcome of this study will be measured using the columns that pertain to HPV vaccination completion. These variables are provider-collected, and they state whether or not the teen is (1) up-to-date with 1+ HPV shot, (2) up-to-date with 2+ HPV shots, or (3) up-to-date with 3+ HPV shots (excluding all vaccinations post-survey). Measuring HPV vaccine completion with “up-to-date” variables may be more robust than vaccination rate based on those who have completed the regime (2 or more shots), because not all teens included in the survey are at the age in which they would’ve completed the full regime. This means that younger teens who are up to date with 1+ shot will not be excluded from the analysis.
  We aim to measure determinants of socioeconomic status that are engrained into the survey questions. These factors include: family income, poverty status, income-to-poverty ratio, insurance status, insurance breaks, maternal education, living arrangement, and geographic mobility status. Additional demographic factors that are associated with healthcare access and may be examined include: race, ethnicity, language, facility in which the vaccine was administered, and whether the teen had completed a wellness exam between the ages of 11-12. Geographic distribution of the teens will be assessed by true state of residence. 
  We would like to examine the data for correlations between HPV vaccination completion status and the aforementioned socioeconomic determinants, demographics, and geographic location. This serves as valuable information to determine differences in healthcare access and vaccination coverage for teenagers living in the U.S.

## Proposed Analysis
	For the analysis portion, we have decided to do a correlation analysis between HPV vaccination completion and each of the socioeconomic factors. When performing this, we may choose to stratify by age or gender, if we find these to be confounders. The correlation analysis cannot be performed using a multivariate model because there may be collinearity present between many of the determinants for socioeconomic status. This will lead to spurious results. Instead, the correlation between vaccination status and each factor is examined separately. 
	Next, we may stratify this by state. Comparing these correlation values between states and regions may lead us to find patterns of healthcare accessibility in the U.S. It is generally known that those of lower income and without insurance have limited healthcare access in the U.S.(1), but it will be interesting to see if this pattern is reflected at the state level. If so, policy should be examined as it may be related to healthcare access disparities. Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 (2). More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic (3). Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.

@@

## References
1. Sciences, N. A. of, Engineering, & Medicine, and. (2018, March 1). Factors that affect health-care utilization. Health-Care Utilization as a Proxy in Disability Determination. https://www.ncbi.nlm.nih.gov/books/NBK500097/ 
2. Mansfield, L. N., Chung, R. J., Silva, S. G., Merwin, E. I., & Gonzalez-Guarda, R. M. (2022). Social determinants of human papillomavirus vaccine series completion among U.S. adolescents: A mixed-methods study. SSM - population health, 18, 101082. https://doi.org/10.1016/j.ssmph.2022.101082
3. Casey, R. M., Akaba, H., Hyde, T. B, et al. (2024). Covid-19 pandemic and equity of global human papillomavirus vaccination: descriptive study of World Health Organization-Unicef vaccination coverage estimates. BMJ Medicine, 3, 000726. doi: 10.1136/bmjmed-2023-000726

>>>>

---
title: "Robertson and Cao Project"
subtitle: "Part 1:Teenage HPV Vaccination Coverage and Socioeconomic Factors"
author: Kelly Cao and Rachel Robertson
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
editor: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Introduction

## Data Set

The data we chose to use is the National Immunization Survey (NIS) of 2022.
It consists survey data that was collected to monitor vaccination coverage for teenagers age 13-17 in the United States of America.
The surveys were conducted by the National Center for Immunization and Respiratory Diseases of the Center (NCIRD) for Disease Control and Prevention (CDC).
The survey itself consists of two parts: (1) the household telephone (random digit dialing) survey answered by a parent or guardian and (2) a mailed survey for the vaccination provider, called “Immunization History Questionnaire.” The original data set is a fixed width file (FWF).
In the data set, there are 43,738 rows and 672 columns, with each row representing an individual and each column representing the answer to a question.
The questions pertain to the teenager’s immunization history, demographics, and additional household-reported health information.
We have truncated the data set to include 26-selected columns for the ease of observation.
We have also filtered the data set to include one year, 2022, which is the most recent year of data that is published.
The provided survey data will require cleaning, as there are signs of human errors present.

## Research Question

We want to use the 2022 National Immunization Survey for teens to identify which socioeconomic and geographic factors are associated with HPV vaccination completion for teenagers in the U.S.
Specifically, we are asking the question, What is the likelihood of a teenager being up-to-date for their HPV vaccinations in the U.S., based on socioeconomic factors and geographic distribution.
The outcome of this study will be measured using the columns that pertain to HPV vaccination completion.
These variables are provider-collected, and they state whether or not the teen is (1) up-to-date with 1+ HPV shot, (2) up-to-date with 2+ HPV shots, or (3) up-to-date with 3+ HPV shots (excluding all vaccinations post-survey).
Measuring HPV vaccine completion with “up-to-date” variables may be more robust than vaccination rate based on those who have completed the regime (2 or more shots), because not all teens included in the survey are at the age in which they would’ve completed the full regime.
This means that younger teens who are up to date with 1+ shot will not be excluded from the analysis.
We aim to measure determinants of socioeconomic status that are engrained into the survey questions.
These factors include: family income, poverty status, income-to-poverty ratio, insurance status, insurance breaks, maternal education, living arrangement, and geographic mobility status.
Additional demographic factors that are associated with healthcare access and may be examined include: race, ethnicity, language, facility in which the vaccine was administered, and whether the teen had completed a wellness exam between the ages of 11-12.
Geographic distribution of the teens will be assessed by true state of residence.
We would like to examine the data for correlations between HPV vaccination completion status and the aforementioned socioeconomic determinants, demographics, and geographic location.
This serves as valuable information to determine differences in healthcare access and vaccination coverage for teenagers living in the U.S.

## Proposed Analysis

```         
For the analysis portion, we have decided to do a correlation analysis between HPV vaccination completion and each of the socioeconomic factors. When performing this, we may choose to stratify by age or gender, if we find these to be confounders. The correlation analysis cannot be performed using a multivariate model because there may be collinearity present between many of the determinants for socioeconomic status. This will lead to spurious results. Instead, the correlation between vaccination status and each factor is examined separately. 
Next, we may stratify this by state. Comparing these correlation values between states and regions may lead us to find patterns of healthcare accessibility in the U.S. It is generally known that those of lower income and without insurance have limited healthcare access in the U.S.(1), but it will be interesting to see if this pattern is reflected at the state level. If so, policy should be examined as it may be related to healthcare access disparities. Recent studies have examined HPV vaccination completion according to social determinants, but they have not examined geographic distribution and include data prior to 2019 (2). More recent years display differences in global HPV vaccination trends due to the COVID19 pandemic (3). Our study will fill in geographical gaps and provide updated vaccination trends based on socioeconomic factors.
```

## Methods

### Data Cleaning and Processing

#### Loading the data

The data was loaded from the original DAT file, entitled 'NISTEENPUF22.DAT' by specifying the column positions within the data file.
The columns were then read and made into a data frame.
Next, each of the factor variables were defined and the corresponding levels were assigned labels.
The structure and summary statistics of all of the variables were examined.
The loaded data is saved as a rds and a csv in the 'processed-data' folder.
This does not alter the original raw data file.
The definition of each factor variable in included in the README file within the 'raw-data' folder.

Next, we explored the structure of the processed data and searched for any missing values.
However, there were no NAs present.
There are, however, missing values as labelled factor levels which must be accounted for.
We identified the missing value labels by printing the factor levels.
We then replaced the all of the values labelled missing with NA values.
Next, we found a string of over 27000 NAs all of the variables for vaccination completion.
These rows corresponded for all variables related to vaccine completion.
Since this is the response variable of our scientific question, we eliminated these rows.
We also eliminated rows with small numbers missing values, which were within the columns for facility, Wellchild exams, and insurance breaks.
Lastly, we eliminatethe year column, as all of the data comes from the year 2022.
We saved the cleaned data as an rds and csv in the 'processed-data' folder.

### Exploratory Analysis

```{r, include=FALSE, cache=FALSE}
library(here)
knitr::read_chunk(here("./code/eda-code/edacode.R"))
```

First we will load the packages needed for the exploratory analysis.
```{r, packages, message = FALSE, warning = FALSE}
```

Next, we will load the data called 'cleandata1' from the 'processed-data' folder.
```{r, loaddata}
```

The following data set has been cleaned.
Any observations with NAs has been removed.

```{r, table1}

@@

The distribution of the numerical income to poverty ratio can be shown below.
It displays a value from 1-3.
A ratio displaying less than 1 depicts an income less than the poverty level.
Any income ratio \>1 indicates an income greater than the poverty level.
Most of the observations were gathered from households with income ratios of 3, as shown by the skewed plot below.

```{r, INCPORAR_I}

The state variable shows a more uniformed distribution, suggesting that the observations were gathered fairly across the US states and territory.
Although it is noted a few territories provides 0 observations.

```{r, STATE}

The family income variable provides a more normal distribution with the majority of the observation coming from \$75000+ income level.

```{r, INCQ298A}

The race and ethnicity variables show a distribution that is similar to the U.S. census data, which concludes that the racial composition of the U.S. is 58.9% white, 13.6% African American, and 19.1% Hispanic or Latino.

```{r, RACEETHK}

```{r, RACE_K}

The Insurance status variable shows that the majority of the observations came from households with private insurance only.
This is followed by any medicaid and then other insurance.
This is mirrored in the facility variable, where a majority of its observations show private facilities being the location.

```{r, INS_STAT2_I}

```{r, FACILITY}

The vaccination variable indicates nearly 80% of the observations had an Up-to-date status for the 1+ shot HPV vaccination.

```{r, Vaccination}

We examine the percentage of up-to-date vaccination status by state.

```{r, state-vaccine}

We examine the facility that submitted the survey, which is stratified by the vaccine completion status of the survey subject.

```{r, facility-vaccine}

We examine the percentage of vaccination completion by race and ethnicity.
Because there are different counts of each race/ethnicity for each factor level, the percentage of those with up-to-date vaccine status out of each race/ethnicity level was determined.
This adjusts for the different counts in each racial category.
The Hispanic ethnicity was found to have the highest vaccine completion while non-hispanic white has the lowest.

```{r, race-ethnicity-vaccine}

### Model and Plot Fitting

A boxplot is created with income-poverty ratio to vaccination status.
The box plot indicates little difference between the two status.
A statistical model is fitted to the same variables.
A p-value of 0.1358 suggests there is little significance between the income-poverty variable and the vaccination status.

```{r, fitfig1}

A boxplot is created with income-poverty ratio to insurance status.
The box plot indicates there are distinct difference between the two status.
A statistical model is fitted to the same variables.
The p-value of \< 2.26e-16 suggests there is statistical significance between the income-poverty variable and the insurance status.

```{r, fitfig2}

Income-poverty ratio was fit to vaccine status and insurance status in a multi-linear model.

```{r, fitfig3}

A box-plot was created for income-poverty level versus race/ethnicity and this was stratified by vaccine completion (UTD or NOT UTD).
A lower income-poverty ratio means the group is closer to the poverty line.
For the non-Hispanic black and Hispanic racial and ethnic groups, vaccine completion corresponded with a lower mean income-poverty ratio.
For mixed ethnic groups, vaccine completion corresponded with a higher mean income-poverty ratio.

```{r, fitfig4}

@@

## References

1.  Sciences, N. A. of, Engineering, & Medicine, and. (2018, March 1). Factors that affect health-care utilization. Health-Care Utilization as a Proxy in Disability Determination. https://www.ncbi.nlm.nih.gov/books/NBK500097/
2.  Mansfield, L. N., Chung, R. J., Silva, S. G., Merwin, E. I., & Gonzalez-Guarda, R. M. (2022). Social determinants of human papillomavirus vaccine series completion among U.S. adolescents: A mixed-methods study. SSM - population health, 18, 101082. https://doi.org/10.1016/j.ssmph.2022.101082
3.  Casey, R. M., Akaba, H., Hyde, T. B, et al. (2024). Covid-19 pandemic and equity of global human papillomavirus vaccination: descriptive study of World Health Organization-Unicef vaccination coverage estimates. BMJ Medicine, 3, 000726. doi: 10.1136/bmjmed-2023-000726
